The Evolution of Phenotypically Invariant Genetic Networks

### 2020-07-06 review-responses.tex

\begin{point}{Eq. 3:}
    What does a coordinate change mean biologically? In our JC, we discussed for a while whether this is just a different parametrization of the exact same biological thing (like transformation to principle components). I think I now understand that it is more, but this should be explained better.
\end{point}

\reply{
    Biologically, a coordinate change may produce an unrecognizably different system
    (repression changed to activation or the like);
    we've added a hopefully clarifying sentence. \revref
}

\begin{point}{}
    ``Since gene networks can grow or shrink following gene duplications and deletions,
these additional degrees of freedom can apply in principle to any system.'' - are there examples of equivalent gene networks with a different size due to gene duplication or loss?
\end{point}

\reply{
    We're not aware of any biological examples of ``equivalent'' networks
    in more than one species at all,
    but we think it's not clear whether we understand the dynamics of \emph{any} gene regulatory network
    well enough to determine equivalence,
    so this is an absence of evidence, not evidence of absence.
    (There are plenty of examples of homologous networks with missing genes,
    but there's so many uncertainties we don't think bringing this up will clarify things.)
    \plr{JOSH: how's this?}
    \jss{Sounds great! We could also add:}
    Over larger evolutionary distances there is some evidence that structurally distinct
    isofunctional networks of different size do indeed exist. \emph{E.g.}, the networks underlying circadian
    rhythm in cyanobacteria and mammals, as discussed by \citet{sancar2008intelligent}.
    It seems possible, in principle, that, on a smaller scale, such differences could exist between more closely related populations.

}

@@

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set N is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

\reply{
    We have removed this part.
}

\begin{point}{(drift along a ridge)}
    In the presence of epistasis, evolution on a neutral network (a high-fitness
  ridge) is *not* due to drift alone, but also affected by (weak second order)
selection in favor of mutational robustness / genetic canalization. In contrast
to what you write, diffusion on the set of network coefficients corresponding
to the optimal phenotype is not unbiased - even if the optimal phenotypes do
indeed all have the same fitness. Instead, selection drives the population to
``thicker'' parts of the network where the mean fitness of the population
(including a cloud of mutants) is higher that on a narrow ridge. This is the
basis of the evolution of robustness/canalization. It is possible to account
for this effect, see Hermisson et al 2003. Amer. Nat. 161, 708-734;
Alvarez-Castro et al, TPB 75 (2009) 109-122, or also Rice, 1998, Evolution 52,
647-656; van Nimwegen, etal 1999, PNAS 96:9716–9720. Note that epistasis is
necessary for the neutral evolution of incompatibilities, which is what you are
aiming for.
\end{point}

\reply{
    \plr{should put somethign about this in the discussion?}
    \jss{I think we should definitely address it in the discussion -- and in the main text of the ridge paper.
    Do you think we can say something about the ridge for the minimal oscillator being constant across parameter space?} 
}

### 2020-11-15 review-responses.tex

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set N is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

\reply{
    We have removed this part.
}

\begin{point}{(drift along a ridge)}
    In the presence of epistasis, evolution on a neutral network (a high-fitness
  ridge) is *not* due to drift alone, but also affected by (weak second order)
selection in favor of mutational robustness / genetic canalization. In contrast
to what you write, diffusion on the set of network coefficients corresponding
to the optimal phenotype is not unbiased - even if the optimal phenotypes do
indeed all have the same fitness. Instead, selection drives the population to
``thicker'' parts of the network where the mean fitness of the population
(including a cloud of mutants) is higher that on a narrow ridge. This is the
basis of the evolution of robustness/canalization. It is possible to account
for this effect, see Hermisson et al 2003. Amer. Nat. 161, 708-734;
Alvarez-Castro et al, TPB 75 (2009) 109-122, or also Rice, 1998, Evolution 52,
647-656; van Nimwegen, etal 1999, PNAS 96:9716–9720. Note that epistasis is
necessary for the neutral evolution of incompatibilities, which is what you are
aiming for.
\end{point}

\reply{
    \plr{should put somethign about this in the discussion?}
    \jss{I think we should definitely address it in the discussion -- and in the main text of the ridge paper.
    Do you think we can say something about the ridge for the minimal oscillator being constant across parameter space?}
    We have added a paragraph to the discussion to address these points. \revref
}

@@

\begin{point}{(drift along a ridge)}
    a second problem results from the fact that evolution on a high-fitness ridge
  often requires coordinated changes at many loci. Take the oscillator system
that you use as an example: simultaneous changes at two genes are required to
maintain the phenotype. This leads to a phenomenon called ``adaptive inertia''
(see Baatz and Wagner, TPB 51, 49-66 and Alvarez-Castro etal . TPB 75, 109-122),
which effectively slows down the movement along the ridge considerably. This
problem applies, in particular, in a ``house-of-cards'' mutation regime when
where rarely two mutations occur together on the same haplotype. In small
populations, it typically requires that populations drift through shallow
fitness valleys. While this is possible, it slows down the process.  
\end{point}

\reply{
    \plr{... together with this?}
    We have added a paragraph to the discussion to address these points. \revref
}

@@

\begin{point}{(drift along a ridge)}
    The relevance of both effects could be studied by simulations in a simple
example (eg the oscillator that is used as an illustration in the ms anyway).
\end{point}

\reply{
    We aim to do that in the subsequent paper.
}

\begin{point}{}
    If speciation due to accumulation of incompatibilities does not occur in
allopatry, but under (even weak) gene flow, some degree of positive selection
is always needed (Bank et al, Genetics 191, 845–863 for a proof). This also
means that even weak gene flow will counteract the process described in the ms.
\end{point}

\reply{
    \plr{... and also this?}
    We have added a paragraph to the discussion to address this point. \revref
}

### 2020-11-15 kalman_walk_paper.tex
fixed figures
fig6
added reference to Example 3 in main text
minor edits to responses
added a discussion paragraph on drift constraits due to canalization, inertia and gene flow.

     \end{tikzpicture} &
     \raisebox{0.25\height}{\includegraphics[scale=0.25]{examples/fig1b}}
\end{tabular}
%  \end{center}
  \caption{
    (Left) 
    Diagram of the gene network in Example \ref{ex:oscillator}
  , and (right) plot of the phenotype $\phi(t)$ against time $t$
  .} \label{fig:oscillator}
\end{figure}

@@

\begin{example}[All equivalent rewirings of the oscillator] \label{ex:all_osc}
The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
by an invertible matrix $V$.
If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
Therefore
the following one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
phenotypically equivalent to the oscillator:
    \begin{align*}
      A(\tau) &= \frac{-1}{\tau+1} \begin{bmatrix} -\tau & -1 \\ 2 \tau(\tau + 1) + 1 &  \tau \end{bmatrix} \ \text{for} \ \tau \neq -1 .
    \end{align*}
The resulting set of systems are depicted in Figure \ref{fig:all_osc}.
\end{example}

@@

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $-2 \tau - \frac{1}{\tau+1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{1}{\tau+1}$} (A); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge[bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[left] {\tiny $\frac{\tau}{\tau+1}$} (A);
    \path [->] (B) edge[loop right] node[right] {\tiny $\frac{-\tau}{\tau+1}$} (B);
\end{scope}
\end{tikzpicture} &	\raisebox{0.25\height}{\includegraphics[scale=0.25]{examples/fig2b}}
    \end{tabular}
% \caption{}
      \caption{
      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (red) differ from $A(-2)$ (green).
      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
      } 
    \label{fig:all_osc}
\end{figure}

@@

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
In practice, we take $\rho(t) = \exp(-t/4\pi)$,
so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
but not longer. \revpoint{AE}{8}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

@@

\begin{figure}[H]
  \centering
 % \begin{tabular}{ccc}
 %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
 % \end{tabular}
  \includegraphics[scale=0.5]{examples/f1f2_newplot}
  \caption{
    \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents;
    parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
    \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
    with parental phenotype again in black.
    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
    Different colored lines correspond to different $F_2$ hybrids;
    note that some completely fail to oscillate.
  } \label{fig:hybs}
\end{figure}

  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
    \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
    $\left(\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2\right)^{1/2}$.
    }
    \label{fig:osc_incompat}
  \end{figure}
\end{example}

@@

>>>>

     \end{tikzpicture} &
     \includegraphics{examples/fig1b_small}
\end{tabular}
%  \end{center}
  \caption{
    (Left) 
    Diagram of the gene network in Example \ref{ex:oscillator}
  , and (right) plot of the phenotype $\phi(t)$ against time $t$
  .} \label{fig:oscillator}
\end{figure}

@@

\begin{example}[All equivalent rewirings of the oscillator] \label{ex:all_osc}
The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
by an invertible matrix $V$.
If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
Therefore
the following one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
phenotypically equivalent to the oscillator:
    \begin{align*}
      A(\tau) &= \frac{-1}{\tau+1} \begin{bmatrix} -\tau & -1 \\ 2 \tau(\tau + 1) + 1 &  \tau \end{bmatrix} \ \text{for} \ \tau \neq -1 .
    \end{align*}
  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. \revpoint{1}{16}
\end{example}

@@

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $-2 \tau - \frac{1}{\tau+1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{1}{\tau+1}$} (A); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge[bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[left] {\tiny $\frac{\tau}{\tau+1}$} (A);
    \path [->] (B) edge[loop right] node[right] {\tiny $\frac{-\tau}{\tau+1}$} (B);
\end{scope}
\end{tikzpicture} &	\includegraphics{examples/fig2b_small}
    \end{tabular}
% \caption{}
      \caption{
      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (red) differ from $A(-2)$ (green).
      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
      } 
    \label{fig:all_osc}
\end{figure}

@@

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
Hybrid phenotypic break down, as a function of genetic distance between
phenotypically equivalent parental oscillators (described in Example \ref{ex:all_osc})
is illustrated in Example \ref{ex:hybrid_osc}.
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
In practice, we take $\rho(t) = \exp(-t/4\pi)$,
so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
but not longer. \revpoint{AE}{8}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

@@

\begin{figure}[H]
  \centering
 % \begin{tabular}{ccc}
 %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
 % \end{tabular}
  %\includegraphics[scale=0.5]{examples/f1f2_newplot}
  \includegraphics{examples/f1f2_small}
  \caption{
    \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents;
    parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
    \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
    with parental phenotype again in black.
    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
    Different colored lines correspond to different $F_2$ hybrids;
    note that some completely fail to oscillate.
  } \label{fig:hybs}
\end{figure}

  \begin{figure}[H]
    \centering
    %\includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
    \includegraphics{examples/fig6_small_maxt10}
    \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
    $\left(\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2\right)^{1/2}$.
    }
    \label{fig:osc_incompat}
  \end{figure}
\end{example}

@@

Phenotypically equivalent system evolution may not solely be driven by drift.
Network coefficient evolution may be biased due to epistasis and driven by second-order selective effects;
such processes can lead an evolving population's exploration of network space to regions more densely populated by
and more closely connected to high fitness system architectures \citep{neutral_network_physics}. Our results, however, do not rely on the ``thickness''
of the phenotypically equivalent set, but on the set's \emph{openess} under sexual recombination.
So long as the phenotypically equivalent set is not closed under sexual recombination,
opportunities for incompatibility remain.
It is plausible, however, that the speed at which system drift can generate incompatibilities might diminish
in circumstances where selection for robustness is strong enough to constrain a population to a small section of system space.
Likewise, as the speed of system drift relies on segregating genetic variation, any constraints on such variation,
possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{alvarez2009effects} or weak gene flow could plausibly slow it down.
\revpoint{1}{18} \revpoint{1}{19} \revpoint{1}{21}
